Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-12 optim.qmd
fill in Section 4

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

# Visualisation of jellyfish optimiser

Information of the jellyfish optimiser is available in a tabular format
and below is an example data collected from finding the sine-wave
structure in 6D data using a distance correlation index (`docr2d_2`):

```{r}
load(here::here("data/sim_example.rda"))
```

```{r}
dplyr::glimpse(sim_example |> select(-id)) 
```

Information recorded can be categorised into the following categories:

-   projection pursuit variables: the index function used (`idx_f`), the
    data dimension (`d`)
-   jellyfish optimiser parameters: the number of jellies (`n_jellies`),
    the maximum number of tries (`max_tries`)
-   simulation variables: the simulation number (`sim`), the seed used
    (`seed`)
-   optimisation variables: the projection basis in a matrix format
    (`basis`), the index value (`index_val`), a description of the
    status - one of "initiation", "current_best", and "jellyfish_update"
    (`info`), current iteration ID (`tries`), current jelly ID (`loop`),
    and the time taken to find the optimum (`time`).

The basis column records every basis *visited* by the jellyfish
optimiser prior to comparing with the current basis. In each iteration,
if the index value of a visited basis is smaller than that of the
current one, the jellyfish optimiser will retain the current basis for
the next iteration, while still documenting the visited basis.

Numerical information to compute:

-   angular distance between the projection basis and the theoretical
    best basis,
-   the proportion of simulation that found the optimal basis,
-   the proportion of jellies, within each simulation, that found the
    optimal basis,

Visualisation to inspect:

-   inspect the basis visited by each jellyfish in the reduced PCA
    space,
-   inspect the final 2D projections reached by each jellyfish,
-   plot the index value against the angular distance between the
    projection basis and the theoretical best basis

Plotting the basis in the space and the projected data can help to
understand 1) whether each simulation finds the same optimum or some
simulations find local optima; and 2) whether the index function used
can detect the structure in the data and the projection contains the
structure of interest.

The visualisations above can be faceted by the projection pursuit
variables and jellyfish optimiser parameters to compare the performance
of different indexes to detect the same structure and how the jellyfish
optimiser parameters affect the optimisation process.

\[example plots\]

```{r}
load(here::here("data/sine_dcor2d_best.rda"))
set.seed(123456)
sin1000 <- spinebil::sinData(6, 1000) %>% scale() %>% as_tibble()
colnames(sin1000) <- paste0("V", 1:6)

proj_dt <- map_dfr(sine_dcor2d_best |> pull(basis), 
              ~as_tibble(as.matrix(sin1000) %*% .x), .id = "sim")

p1 <- proj_dt |>
  ggplot(aes(x = V1, y = V2)) + 
  geom_point(size = 0.1) + 
  xlim(-4, 4) + ylim(-4, 4) +
  facet_wrap(~as.numeric(sim), ncol = 10) + 
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
```

```{r}
#| fig.height: 5 
#| fig.width: 10
#| fig.cap: sdfjsflk
p1 
```

```{r eval = FALSE}
theo_best <- matrix(c(rep(0, 8), 1, 0, 0, 1), nrow = 6, byrow = TRUE)
angle_dcor2d <- sine_dcor2d_5050 |>
    rowwise() |>
    mutate(angle = tourr::proj_dist(theo_best, basis))

p2 <- angle_dcor2d |>
  ggplot(aes(x = angle, y = index_val)) +
  geom_point(size = 0.1) +
  theme(aspect.ratio = 1)

@@

In the second experiment, fifty (50) simulations are conducted at each  hyper-parameter combination to analyse its effects on the success rate of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and 100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each simulation,  success is defined as achieving a final index value within 0.05 of the best index value found among all 50 simulations. For example, in the 8-dimensional data scenario with 20 jellyfishes and a maximum number of tries set to 100, the optimal index value is `r pipe_sim_best |> filter(n_jellies == 20, max_tries == 100, d == 8) |> pull(I_best) |> round(3) -> aa; aa`. Out of the 50 simulations, `r pipe_run_summ |> filter(n_jellies == 20, max_tries == 100, d == 8) |> filter(I_max >= aa - 0.05) |> nrow() -> bb; bb` achieved an index value within [`r aa - 0.05`, `r aa`], resulting in a success rate of `r pipe_run_summ |> filter(n_jellies == 20, max_tries == 100, d == 8) |> mutate(a = I_max >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`. @fig-proportion presents the proportion of success for each hyper-parameter combinations. As the number of jellyfish and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions), small parameter values (20 jellyfishes and a max number of tries of 50) can already result in high success rate, while for higher-dimensional problems (8, 10, and 12 dimensions), a combination of 100 jellyfishes and 100 maximum number of tries is necessary for at least half of the jellyfishes to find the optimal projection.

@@

*Note: should we print out the 52 cases investigated somewhere, say appendex, since it is not full grid simulation (otherwise for the sine-wave problem there will be 6 indexes x 6 configurations = 36 cases for 6-D data and 2 indexes [MIC & TIC] x 6 configurations = 12 cases for 8-D data)? We only do a selection of 28 of them to save time/ computational resources.*

@@

@tbl-mod-data combines all the variables calculated above (the JS optimiser success rate, jellyfish hyper-parameters, and optimisation properties) into one table. Three pre-processing steps are applied to the data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for interpretation, 2) creating a new binary variable `long_time` to indicate cases with an average run time over 30 seconds, and 3) encoding the success rate for the stringy case as 0, since none of the 50 simulations identified the sine-wave shape. A generalised linear model with a binomial family and a logit link function is used to fit the data and @tbl-mod-output presents the model outputs. The model suggests that the success rate of the JS optimiser is positively associated with the two jellyfish hyper-parameters, as well as with smoothness and squintability. However, being flagged with long runtime and an increase of data dimension will reduce the success rate. The variable `squintability` and `dimension` are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation sucess.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum numer of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with teh success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>  
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J_hat = ifelse(index == "stringy", 0, P_J_hat))

>>>>

```{r setup, echo = FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggh4x)
library(broom)
library(kableExtra)
library(ferrn)
load(here::here("data/sim_pipe.rda"))
load(here::here("data/sim_sine_6d_spline_head.rda"))
load(here::here("data/sim_sine_6d_spline_best.rda"))
load(here::here("data/sim_sine_6d_spline_projdist.rda"))
load(here::here("data/pipe_better.rda"))
load(here::here("data/pipe_jellyfish.rda"))
load(here::here("data/smoothness.rda"))
load(here::here("data/squintability.rda"))
load(here::here("data/sq_basis_dist_idx.rda"))
load(here::here("data/sim_df.rda"))
```

@@

# Visualisation of jellyfish optimiser {#sec-vis}

The data structure proposed in @RJ-2021-105 is implemented for JS optimisers and  additional information such as index function used, data dimension, and jellyfish hyper-parameters can be included as additional columns. Below prints an example data collected from finding the sine-wave structure in 6D data (`d = 6`) using a spline index function(`splines2d`):

```{r}
dplyr::glimpse(sim_sine_6d_spline_head |> mutate(index = "spline2d") |> select(-id, -info)) 
```

The jellyfish hyper-parameters used are 50 jellyfishes (`n_jellies`) and a maximum of 50 tries (`max_tries`). The full data contains 50 simulations, recorded by the variable `sim`, with different seeds (`seed`). Recorded in columns `basis`, `index_val`, `tries`, `loop` and `time`, are the sampled projection basis, calculated index value, iteration ID, jellyfish ID, and time taken to find the optimum, respectively. The basis column records every visited basis before comparing with the current (last) basis. If a visited basis has a smaller index value than the current (last) one, the JS optimiser keeps the current basis for the next iteration in the search but still records the visited inferior basis. This data can be used to derive numerical summaries and plots:

-   **success rate**:  Success rate is defined as the proportion of simulations that achieve a final index value within 0.05 of the best index value found among all 50 simulations. In the dataset above, the optimal index value is `r sim_sine_6d_spline_best |> pull(index_val) |> max() |> round(3) -> aa; aa`. Out of the 50 simulations, `r sim_sine_6d_spline_best |> filter(index_val >= aa - 0.05) |> nrow() -> bb; bb` achieved an index value within [`r aa - 0.05`, `r aa`], resulting in a success rate of `r sim_sine_6d_spline_best |> mutate(a = index_val >= aa - 0.05) |> pull(a) -> cc; round(sum(cc)/length(cc), 3)`. @fig-projection shows the best projection found in each simulation, sorted by the index value. Most of the projections show a clear sine-wave structure, while the last six shows only a straight line with clusters at the end. When local optima exist in the projection pursuit problem, plotting the projections found in each simulation or by each jellyfish can help identify these local optima.

-  **Projection distance**: The projection distance between projection bases and the theoretical best basis (or the empirical best basis if the theoretical best is unknown) is calculated as the Frobenius norm, which measures how close a visited basis is to the best basis. Plotting this projection distance against the index value for each basis visited reveals how the index value changes as the basis approaches the optimal basis, helping to identify the squintability of the index function.  @fig-index-value-proj-dist shows the index value against the projection distance for the given example and its binned version. As the projection distance decreases, the index value initially increases with an increasing slope and then level off. The index value shows greater variation at medium projection distances compared to small and large distances.

```{r}
#| label: fig-projection
#| fig.height: 6
#| fig.width: 10
#| fig.cap: Best projections found by the JS opitmiser in each simulation, sorted by the index value. The projection pursuit problem is to detect the 6D sine-wave structure using the spline index. Most simulations have detected a clear sine-wave pattern, while the last six show a straight line with clustering at the ends.. In this simulation, the success rate is 0.88 (44/50).
set.seed(123456)
sin1000 <- spinebil::sinData(6, 1000) %>% scale() %>% as_tibble()
colnames(sin1000) <- paste0("V", 1:6)

proj_dt <- map_dfr(sim_sine_6d_spline_best |> pull(basis), 
              ~as_tibble(as.matrix(sin1000) %*% .x), .id = "sim") |> 
  mutate(sim = as.numeric(sim)) |> 
  left_join(sim_sine_6d_spline_best |> select(sim, index_val), by = "sim")

proj_dt |>
  ggplot(aes(x = V1, y = V2)) + 
  geom_point(size = 0.1) + 
  geom_text(aes(label = round(index_val, 3)), x = 0, y = 3.7, 
            size = 3) +
  xlim(-4, 4) + ylim(-4, 4) +
  facet_wrap(~fct_reorder(as.factor(sim), -index_val), ncol = 10) + 
  theme_bw() +
  theme(aspect.ratio = 1, axis.ticks = element_blank(),
        axis.text = element_blank(), axis.title = element_blank(),
        panel.grid = element_blank())
```

```{r}
#| label: fig-index-value-proj-dist
#| fig.height: 3
#| fig.width: 8
#| fig.cap: "Projection distance plotted against the index value for detecting the 6D sine-wave structure using the spline index function. a) The index values start around 0 with low variance at large projection distances. As the projection distance decreases, the index values increase with a larger variance. Towards the minimum projection distance, the index values increase and cluster around 1, indicating the optimiser is approaching to the optimum. b)  Projection distances are binned with a bin width of 0.1, and index values are averaged over each bin. The relationship between the two variables forms a sigmoid curve."
p1 <- sim_sine_6d_spline_projdist |> 
  ggplot(aes(x = proj_dist, y = index_val)) +
  geom_point(size = 0.1, alpha = 0.5) + 
  labs(x = "Projection distance", y = "Index value")

p2 <- sim_sine_6d_spline_projdist |> 
  group_by(proj_bin = proj_dist %/% 0.1) |> 
  summarise(index_val = mean(index_val)) |> 
  ggplot(aes(x = proj_bin, y = index_val)) + 
  geom_line() +
  geom_point() + 
  labs(x = "Projection distance", y = "Index value")

(p1 | p2) + plot_annotation(tag_levels = "a")
```

@@

In the second experiment, fifty (50) simulations are conducted at each  hyper-parameter combination to analyse its effects on the success rate of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and 100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each simulation,  success rate is calculated as defined in @sec-vis. @fig-proportion presents the proportion of success for each hyper-parameter combinations. As the number of jellyfish and maximum tries increase, the success rate also increases. For simpler problems (6 dimensions), small parameter values (20 jellyfishes and a maximum number of tries of 50) can already result in high success rate, while for higher-dimensional problems (8, 10, and 12 dimensions), a combination of 100 jellyfishes and 100 maximum number of tries is necessary for at least half of the jellyfishes to find the optimal projection.

@@

*Note: should we print out the 52 cases investigated somewhere, say appendix, since it is not full grid simulation (otherwise for the sine-wave problem there will be 6 indexes x 6 configurations = 36 cases for 6-D data and 2 indexes [MIC & TIC] x 6 configurations = 12 cases for 8-D data)? We only do a selection of 28 of them to save time/ computational resources.*

@@

@tbl-mod-data combines all the variables calculated above (the JS optimiser success rate, jellyfish hyper-parameters, and optimisation properties) into one table. Three pre-processing steps are applied to the data: 1) scaling the jellyfish hyper-parameters by a factor of 10 for interpretation, 2) creating a new binary variable `long_time` to indicate cases with an average run time over 30 seconds, and 3) encoding the success rate for the stringy case as 0, since none of the 50 simulations identified the sine-wave shape. A generalised linear model with a binomial family and a logit link function is used to fit the data and @tbl-mod-output presents the model outputs. The model suggests that the success rate of the JS optimiser is positively associated with the two jellyfish hyper-parameters, as well as with smoothness and squintability. However, being flagged with long runtime and an increase of data dimension will reduce the success rate. The variable `squintability` and `dimension` are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>  
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J_hat = ifelse(index == "stringy", 0, P_J_hat))

### 2024-06-12 optim.tex
fill in Section 4

\begin{tcolorbox}[enhanced jigsaw, toprule=.15mm, coltitle=black, rightrule=.15mm, titlerule=0mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, left=2mm, breakable, colbacktitle=quarto-callout-note-color!10!white, colback=white, toptitle=1mm, opacityback=0, opacitybacktitle=0.6, arc=.35mm, bottomtitle=1mm, leftrule=.75mm, colframe=quarto-callout-note-color-frame]

@@

\section{Visualisation of jellyfish
optimiser}\label{visualisation-of-jellyfish-optimiser}

Information of the jellyfish optimiser is available in a tabular format
and below is an example data collected from finding the sine-wave
structure in 6D data using a distance correlation index
(\texttt{docr2d\_2}):

\begin{verbatim}
Rows: 10
Columns: 13
$ idx_f     <chr> "dcor2d_2", "dcor2d_2", "dcor2d_2", "dcor2d_2", "dcor2d_2", ~
$ d         <dbl> 6, 6, 6, 6, 6, 6, 6, 6, 6, 6
$ n_jellies <dbl> 20, 20, 20, 20, 20, 20, 20, 20, 20, 20
$ max_tries <dbl> 50, 50, 50, 50, 50, 50, 50, 50, 50, 50
$ sim       <int> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1
$ seed      <int> 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462
$ basis     <list> <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<m~
$ index_val <dbl> 0.0247212373, 0.0033938502, 0.0463398915, 0.0486230801, -0.0~
$ info      <chr> "initiation", "initiation", "initiation", "initiation", "ini~
$ method    <chr> "search_jellyfish", "search_jellyfish", "search_jellyfish", ~
$ tries     <dbl> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1
$ loop      <dbl> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10
$ time      <drtn> 35.46031 secs, 35.46031 secs, 35.46031 secs, 35.46031 secs, ~
\end{verbatim}

Information recorded can be categorised into the following categories:

\begin{itemize}
\tightlist
\item
  projection pursuit variables: the index function used
  (\texttt{idx\_f}), the data dimension (\texttt{d})
\item
  jellyfish optimiser parameters: the number of jellies
  (\texttt{n\_jellies}), the maximum number of tries
  (\texttt{max\_tries})
\item
  simulation variables: the simulation number (\texttt{sim}), the seed
  used (\texttt{seed})
\item
  optimisation variables: the projection basis in a matrix format
  (\texttt{basis}), the index value (\texttt{index\_val}), a description
  of the status - one of ``initiation'', ``current\_best'', and
  ``jellyfish\_update'' (\texttt{info}), current iteration ID
  (\texttt{tries}), current jelly ID (\texttt{loop}), and the time taken
  to find the optimum (\texttt{time}).
\end{itemize}

The basis column records every basis \emph{visited} by the jellyfish
optimiser prior to comparing with the current basis. In each iteration,
if the index value of a visited basis is smaller than that of the
current one, the jellyfish optimiser will retain the current basis for
the next iteration, while still documenting the visited basis.

Numerical information to compute:

\begin{itemize}
\tightlist
\item
  angular distance between the projection basis and the theoretical best
  basis,
\item
  the proportion of simulation that found the optimal basis,
\item
  the proportion of jellies, within each simulation, that found the
  optimal basis,
\end{itemize}

Visualisation to inspect:

\begin{itemize}
\tightlist
\item
  inspect the basis visited by each jellyfish in the reduced PCA space,
\item
  inspect the final 2D projections reached by each jellyfish,
\item
  plot the index value against the angular distance between the
  projection basis and the theoretical best basis
\end{itemize}

Plotting the basis in the space and the projected data can help to
understand 1) whether each simulation finds the same optimum or some
simulations find local optima; and 2) whether the index function used
can detect the structure in the data and the projection contains the
structure of interest.

The visualisations above can be faceted by the projection pursuit
variables and jellyfish optimiser parameters to compare the performance
of different indexes to detect the same structure and how the jellyfish
optimiser parameters affect the optimisation process.

{[}example plots{]}

\begin{figure}[H]

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-4-1.pdf}

@@

\caption{sdfjsflk}

@@

In the second experiment, fifty (50) simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, success is defined as achieving a final index value within
0.05 of the best index value found among all 50 simulations. For
example, in the 8-dimensional data scenario with 20 jellyfishes and a
maximum number of tries set to 100, the optimal index value is 0.997.
Out of the 50 simulations, 27 achieved an index value within {[}0.947,
0.997{]}, resulting in a success rate of 0.54.
Figure~\ref{fig-proportion} presents the proportion of success for each
hyper-parameter combinations. As the number of jellyfish and maximum
tries increase, the success rate also increases. For simpler problems (6
dimensions), small parameter values (20 jellyfishes and a max number of
tries of 50) can already result in high success rate, while for
higher-dimensional problems (8, 10, and 12 dimensions), a combination of
100 jellyfishes and 100 maximum number of tries is necessary for at
least half of the jellyfishes to find the optimal projection.

@@

\emph{Note: should we print out the 52 cases investigated somewhere, say
appendex, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes {[}MIC \& TIC{]} x 6 configurations = 12
cases for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.}

@@

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three pre-processing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of the JS
optimiser is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to jellyfish hyper-parameters in the optimisation sucess.

@@

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum numer of tries are
positively associated with the JS optimiser success rate while data
dimension and being flagged as long runtime are negatively associated
with teh success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.}

>>>>

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, left=2mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, toptitle=1mm, colback=white, colframe=quarto-callout-note-color-frame, titlerule=0mm, breakable, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, bottomrule=.15mm, arc=.35mm, rightrule=.15mm, leftrule=.75mm, opacityback=0]

@@

\section{Visualisation of jellyfish optimiser}\label{sec-vis}

The data structure proposed in \citet{RJ-2021-105} is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. Below prints an example data collected from finding the
sine-wave structure in 6D data (\texttt{d\ =\ 6}) using a spline index
function(\texttt{splines2d}):

\begin{verbatim}
Rows: 10
Columns: 12
$ n_jellies <dbl> 50, 50, 50, 50, 50, 50, 50, 50, 50, 50
$ max_tries <dbl> 50, 50, 50, 50, 50, 50, 50, 50, 50, 50
$ d         <dbl> 6, 6, 6, 6, 6, 6, 6, 6, 6, 6
$ sim       <int> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1
$ seed      <int> 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462, 3462
$ basis     <list> <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<matrix[6 x 2]>>, <<m~
$ index_val <dbl> 0.036664532, 0.005121912, 0.058683104, 0.056527991, 0.015734~
$ method    <chr> "search_jellyfish", "search_jellyfish", "search_jellyfish", ~
$ tries     <dbl> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1
$ loop      <dbl> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10
$ time      <drtn> 4.362133 mins, 4.362133 mins, 4.362133 mins, 4.362133 mins, ~
$ index     <chr> "spline2d", "spline2d", "spline2d", "spline2d", "spline2d", ~
\end{verbatim}

The jellyfish hyper-parameters used are 50 jellyfishes
(\texttt{n\_jellies}) and a maximum of 50 tries (\texttt{max\_tries}).
The full data contains 50 simulations, recorded by the variable
\texttt{sim}, with different seeds (\texttt{seed}). Recorded in columns
\texttt{basis}, \texttt{index\_val}, \texttt{tries}, \texttt{loop} and
\texttt{time}, are the sampled projection basis, calculated index value,
iteration ID, jellyfish ID, and time taken to find the optimum,
respectively. The basis column records every visited basis before
comparing with the current (last) basis. If a visited basis has a
smaller index value than the current (last) one, the JS optimiser keeps
the current basis for the next iteration in the search but still records
the visited inferior basis. This data can be used to derive numerical
summaries and plots:

\begin{itemize}
\item
  \textbf{success rate}: Success rate is defined as the proportion of
  simulations that achieve a final index value within 0.05 of the best
  index value found among all 50 simulations. In the dataset above, the
  optimal index value is 0.999. Out of the 50 simulations, 44 achieved
  an index value within {[}0.949, 0.999{]}, resulting in a success rate
  of 0.88. Figure~\ref{fig-projection} shows the best projection found
  in each simulation, sorted by the index value. Most of the projections
  show a clear sine-wave structure, while the last six shows only a
  straight line with clusters at the end. When local optima exist in the
  projection pursuit problem, plotting the projections found in each
  simulation or by each jellyfish can help identify these local optima.
\item
  \textbf{Projection distance}: The projection distance between
  projection bases and the theoretical best basis (or the empirical best
  basis if the theoretical best is unknown) is calculated as the
  Frobenius norm, which measures how close a visited basis is to the
  best basis. Plotting this projection distance against the index value
  for each basis visited reveals how the index value changes as the
  basis approaches the optimal basis, helping to identify the
  squintability of the index function.
  Figure~\ref{fig-index-value-proj-dist} shows the index value against
  the projection distance for the given example and its binned version.
  As the projection distance decreases, the index value initially
  increases with an increasing slope and then level off. The index value
  shows greater variation at medium projection distances compared to
  small and large distances.
\end{itemize}

\includegraphics{optim_files/figure-pdf/fig-projection-1.pdf}

@@

\caption{\label{fig-projection}Best projections found by the JS
opitmiser in each simulation, sorted by the index value. The projection
pursuit problem is to detect the 6D sine-wave structure using the spline
index. Most simulations have detected a clear sine-wave pattern, while
the last six show a straight line with clustering at the ends.. In this
simulation, the success rate is 0.88 (44/50).}

\includegraphics{optim_files/figure-pdf/fig-index-value-proj-dist-1.pdf}

\caption{\label{fig-index-value-proj-dist}Projection distance plotted
against the index value for detecting the 6D sine-wave structure using
the spline index function. a) The index values start around 0 with low
variance at large projection distances. As the projection distance
decreases, the index values increase with a larger variance. Towards the
minimum projection distance, the index values increase and cluster
around 1, indicating the optimiser is approaching to the optimum. b)
Projection distances are binned with a bin width of 0.1, and index
values are averaged over each bin. The relationship between the two
variables forms a sigmoid curve.}

@@

In the second experiment, fifty (50) simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, success rate is calculated as defined in
Section~\ref{sec-vis}. Figure~\ref{fig-proportion} presents the
proportion of success for each hyper-parameter combinations. As the
number of jellyfish and maximum tries increase, the success rate also
increases. For simpler problems (6 dimensions), small parameter values
(20 jellyfishes and a maximum number of tries of 50) can already result
in high success rate, while for higher-dimensional problems (8, 10, and
12 dimensions), a combination of 100 jellyfishes and 100 maximum number
of tries is necessary for at least half of the jellyfishes to find the
optimal projection.

@@

\emph{Note: should we print out the 52 cases investigated somewhere, say
appendix, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes {[}MIC \& TIC{]} x 6 configurations = 12
cases for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.}

@@

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three pre-processing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of the JS
optimiser is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to jellyfish hyper-parameters in the optimisation success.

@@

\caption{\label{tbl-mod-output}Model estimates of proportion of
jellyfish success on optimisation properties and jellyfish
hyper-parameters from the generalised linear model with a binomial
family and a logit link function. The variable smoothness,
squintability, number of jellyfish and maximum number of tries are
positively associated with the JS optimiser success rate while data
dimension and being flagged as long runtime are negatively associated
with the success rate. The variable squintability and dimension are
significant, suggesting their importance relative to jellyfish
hyper-parameters in the optimisation success.}
